---
title: "Chapter 2: Generalized Linear Mixed Models"
author: "Aritz Adin y Jaione Etxeberria"
date: "02-01-2024"
date-format: "MMM/YYYY"
format:
  html:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)  
```

# Exercise 1

Data from a clinical trial, involving 2 treatments (control and active drug), conducted at 11 randomly selected centres are discussed by [Beitler and Landis (1985)](https://www.jstor.org/stable/2530970). For the $i$-th center and $j$-th treatment, the proportion of $n_{ij}$ patients having a positive response ($y_{ij}$) is recorded below (*material extracted from Gallagher, 2023*):

| **Centre** | **Treatment** | $y_{ij}$ | $n_{ij}$ |
|:----------:|:-------------:|:--------:|:--------:|
|     1      |     drug      |    11    |    36    |
|     1      |    control    |    10    |    37    |
|     2      |     drug      |    16    |    20    |
|     2      |    control    |    22    |    32    |
|     3      |     drug      |    14    |    19    |
|     3      |    control    |    7     |    19    |
|  $\ldots$  |   $\ldots$    | $\ldots$ | $\ldots$ |

### 1. Load the data intored in the `multicentre.txt` file, convert `centre` variables to factor, and make a descriptive plot to visualize differences between treatments and/or centers.

```{r}
multicentre <- read.table(file="multicentre.txt", header=T, stringsAsFactors=T)
head(multicentre)

multicentre$centre <- as.factor(multicentre$centre)
str(multicentre)
```

We can make a descriptive plot of the data using the ggplot2 package

```{r}
library(ggplot2)

library(ggplot2)
ggplot(multicentre, aes(x=obs, y=centre, color=treat)) +
  geom_point() +
  labs(x="Observed probability", y="Centre")
```

We observe systematic differences between centres.

### 2. Which variable should be included as a fixed effect, and which as a random effect? Make a *design plot* to visually compare the magnitude of the effects of the `centre` and `treat` factors.

-   We want to compare these particular treatments (*experimental factor*), so we use fixed effects for the `treat` factor.

-   The eleven centres represents a random sample from the population about which we wish to make inferences (*random factor*), so we use random effects to model the `center` factor.

```{r, fig.cap="Average observed probabilities for each level of the factors `treat` and `center`"}
plot.design(obs ~ treat*centre, data=multicentre)
```

-   We see that the variability associated with the treatment is much lower than the variability associated with the centres.

-   We also see that the average probability of a positive response is higher in the *active drug* treatment.

### 3. Write the mixed logistic regression model equation.

$$y_{ij} | u_i  \sim Bin(n_{ij},\pi_{ij}), \quad i=1,\ldots,10 \quad \mbox{and} \quad j=1,2$$ $$\mbox{logit}(\pi_{ij}) = \beta_0 + \beta_1*x_{ij} + u_i, \quad u_i \sim N(0,\sigma_u^2)$$ where

-   $\pi_{ij}$ is the probability of a favourable outcome in a patient on the $i$-th centre and $j$-th treatment
-   $x_{ij}$ is an indicator variable for treatment (0=control, 1=active drug)
-   $u_i$ is a random variable associated with the $i$-th centre


### 4. Test if random effects are necessary.

```{r, message=FALSE, include=TRUE}
## Fit the models ##
library(lme4)

M0 <- glm(obs ~ 1 + treat, family="binomial", weights=n, data=multicentre)
M1 <- glmer(obs ~ 1 + treat + (1|centre), family="binomial", weights=n, data=multicentre)

## LRT for the variance component sigma2_u ##
LRT <- -2*(logLik(M0)-logLik(M1))
mean(1-pchisq(LRT,df=c(0,1)))

## We can also use the AIC/BIC to compare the models ##
data.frame(AIC=c(AIC(M0), AIC(M1)),
           BIC=c(BIC(M0), BIC(M1)),
           row.names=c("multicentre.glm","multicentre.glmer"))
```

### 5. Check if the treatment effect is significant.

We compare nested models (with and without `treat` factor) using the LRT with the `anova()` function

```{r, include=TRUE}
M1a <- glmer(obs ~ 1 + (1|centre), family="binomial", weights=n, data=multicentre)
M1b <- glmer(obs ~ 1 + treat + (1|centre), family="binomial", weights=n, data=multicentre)
anova(M1a,M1b)
```
**We conclude that the treatment effect is not statistically significant.**


### 6. Using the model with fixed and random effects, which is the probability of a positive response on the active drug group? Interpret the odds ratio $e^{\beta_1}$ and compute a 95% confidence interval.

```{r, include=TRUE}
library(emmeans)
emmeans(M1b, revpairwise~treat, type="response", inf=TRUE)
```

-  For $u_i=0$, the probability of a positive response over a population of centres under the active drug group ($x_{ij}=1$) is $\hat{\pi}_{ij}=0.376$ with a 95% confidence interval of $[0.220, 0.564]$.

-   The estimated odds of a patient showing a positive response on the active drug group relative to the control group is 1.08 with a 95% confidence interval of $[0.77, 1.51]$.


### 7. Using the best model, verify whether the model assumptions on the random effects are satisfied.

```{r, include=TRUE}
Model <- M1a

rand <- unlist(ranef(Model)$centre)
qqnorm(rand)
qqline(rand)

shapiro.test(rand)
```
We don't reject the null hypothesis of normality of the random effects.

### 8. Which is the estimated variance and 95% confidence interval of the within-centre random effect? Compute and interpret the intra-class correlation coefficient.

```{r, include=TRUE}
VarCorr(Model)
confint(Model, parm="theta_")
```

We can compute the intra-class correlation coefficient as $$\rho=\frac{\sigma^2_u}{\sigma^2_u+3.29}=1.169^2/(1.169^2+3.29) = 0.29347,$$ which means that 29% of the overall variability can be attributed to the differences between centres.

### 9. Compute the predicted probabilties of a positive response for each value of the dataset. Include those estimated probabilities in the descriptive plot of section 1).

```{r, fig.cap="Observed (dots) and predicted (asterisks) probabilities of a positive response for each centre"}
multicentre$pred <- predict(Model, newdata=multicentre, type="response")
head(multicentre)

ggplot(multicentre, aes(x=obs, y=centre, color=treat)) +
  geom_point() +
  geom_point(aes(x=pred, y=centre), shape=8, color="black") + 
  labs(x="Prob", y="Centre")
```

# Exercise 2

A Health Maintenance Organization wants to know what patient and physician factors are most related to whether a patient’s lung cancer goes into remission after treatment as part of a larger study of treatment outcomes and quality of life in patients with lung cancer.

The data has a multilevel structure where 8525 patients are nested within 407 doctors, who are in turn nested within 35 hospitals. The dataset contains either individual-level, doctor-level and hospital-level explanatory variables:

* `remission`: response variable (0=no, 1=yes)
* `age`: age of the patient (in years)
* `IL6`: Interleukin-6 concentration in blood (pg/ml)
* `CRP`: C-reactive protein concentration in blood (mg/dl)
* `LengthofStay`: duration of a patient's hospital stay (in weeks)
* `CancerStage`: lung cancer stage (I, II, III or IV)
* `Smoking`: is the patient a smoker? (1=never, 2=former, 3=current)
* `ID.doctor`: ID of the doctor
* `ID.hospital`: ID of the hospital


### 1. Load the data intored in the `Patients.txt` file. Convert `ID.doctor` and `ID.hospital` variables to factor. 

```{r}
patients <- read.table(file="patients.txt", header=T, stringsAsFactors=T)
head(patients)

patients$ID.doctor <- as.factor(patients$ID.doctor)
patients$ID.hospital <- as.factor(patients$ID.hospital)
patients$Smoking <- factor(patients$Smoking, levels=c("never","former","current"))
str(patients)
```
### 2. Make some descriptive plots to visualize the relationship between variables.

```{r, warning=FALSE, message=FALSE}
library(ggplot2)

## Explanatory continuous variables ##
plot(patients[, c("Age","IL6","CRP","LengthofStay")])

## Explanatory categorical variables ##
ggplot(patients, aes(x=CancerStage, y=LengthofStay)) +
  stat_sum(aes(size=after_stat(n), group=1)) +
  scale_size_area(max_size=10)

## Response variable change over explanatory variables ##
tmp <- reshape2::melt(patients[, c("remission", "IL6", "CRP")], id.vars="remission")
ggplot(tmp, aes(factor(remission), y = value, fill=factor(remission))) +
  geom_boxplot() +
  facet_wrap(~variable, scales="free_y")

ggplot(patients, aes(x=Age, y=remission, color=Smoking)) +
  geom_point() +
  stat_smooth(method="glm", color="black", se=FALSE,
              method.args = list(family=binomial)) +
  ggtitle("Lung cancer remission after treatment by age and cancer stage") + 
  facet_wrap(~ CancerStage)
```

### 3. Test if random effects are necessary.

<!-- For $i=1,\ldots,407$ (doctors) and $k=1,\ldots,35$ (hospitals): -->

<!-- * Model 0: $\quad \mbox{logit}(\pi_{ijk})=\beta_0$ -->

<!-- * Model 1: $\quad \mbox{logit}(\pi_{ijk})=\beta_0 + u_i, \quad u_i \sim N(0,\sigma^2_u)$ -->

<!-- * Model 2: $\quad \mbox{logit}(\pi_{ijk})=\beta_0 + u_i + v_j, \quad u_i \sim N(0,\sigma^2_u), \quad v_j \sim N(0,\sigma^2_u)$ -->

<!-- where $\pi_{ijk}$ is the probability of a remission in lung cancer in a patient attended by the $i$-th doctor within the $j$-th hospital, $u_i$ is a doctor-level random effect and $v_k$ is a hospital-level random effect. -->

```{r, message=FALSE, include=FALSE}
library(lme4)

# Logistic regression model without random effects #
M0 <- glm(remission ~ 1, family="binomial", data=patients)

# Logistic regression model with doctor-level random effects #
M1 <- glmer(remission ~ 1 + (1|ID.doctor), family="binomial", data=patients)

# Logistic regression model with doctor-level and hospital-level random effects #
M2 <- glmer(remission ~ 1 + (1|ID.doctor) + (1|ID.hospital),
            family="binomial", data=patients)

## LRT for the variance components ##
LRT1 <- -2*(logLik(M0)-logLik(M1))
mean(1-pchisq(LRT1,df=c(0,1)))

LRT2 <- -2*(logLik(M1)-logLik(M2))
mean(1-pchisq(LRT2,df=c(0,1)))

## Model comparison with AIC/BIC ##
data.frame(AIC=c(AIC(M0), AIC(M1), AIC(M2)),
           BIC=c(BIC(M0), BIC(M1), BIC(M2)),
           row.names=c("glm","glmer.1","glmer.2"))
```

### 4. Hypothesis test for the fixed effects

```{r, message=FALSE, include=FALSE}
## Model with random effects + continuous predictors ##
M3a <- glmer(remission ~ 1 + Age + IL6 + CRP + LengthofStay + (1|ID.doctor) + (1|ID.hospital),
             family="binomial", data=patients)
summary(M3a)
anova(M2,M3a)

## Model with random effects + continuous predictors + CancerStage ##
M3b <- glmer(remission ~ 1 + Age + IL6 + CRP + LengthofStay + CancerStage + (1|ID.doctor) + (1|ID.hospital),
             family="binomial", data=patients)
summary(M3b)
anova(M3a,M3b)

M3c <- glmer(remission ~ 1 + IL6 + CRP + LengthofStay + CancerStage + (1|ID.doctor) + (1|ID.hospital),
             family="binomial", data=patients)
summary(M3c)
anova(M3b,M3c)

## Model with random effects + continuous predictors + CancerStage + Smoking ##
M3d <- glmer(remission ~ 1 + IL6 + CRP + LengthofStay + CancerStage + Smoking + (1|ID.doctor) + (1|ID.hospital),
             family="binomial", data=patients)
summary(M3d)
anova(M3b,M3d)
```

### 5. Fit the final logistic mixed model and write its equation. Compute intra-class correlation coefficient(s).

<!-- $$\mbox{logit}(\pi_{ijk})= 0.55 - 0.058*IL6_{ijk} - 0.023*CRP_{ijk} - 0.152*Length_{ijk} - 0.55*I(CancerStage="II")_{ijk} - 1.214*I(CancerStage="III")_{ijk} - 2.581*I(CancerStage="IV")_{ijk} - 0.379*I(Smoking="former")_{ijk} - 0.465*I(Smoking="current")_{ijk}$$ -->

```{r, message=FALSE, include=FALSE}
Model.final <- M3d
fixef(Model.final)

VarCorr(Model.final)
ICC.doctor <- 1.989^2/(1.989^2+0.471^2+3.29)
ICC.doctor

ICC.hospital <- 0.471^2/(1.989^2+0.471^2+3.29)
ICC.hospital
```

### 6. Verify wether the model assumptions are satisfied

```{r, message=FALSE, include=FALSE}
rand.u <- unlist(ranef(Model.final)$ID.doctor)
qqnorm(rand.u)
qqline(rand.u)
shapiro.test(rand.u)

rand.v <- unlist(ranef(Model.final)$ID.hospital)
qqnorm(rand.v)
qqline(rand.v)
shapiro.test(rand.v)
```


### 7. Examine the model results
```{r, message=FALSE, include=FALSE}
library(emmeans)

emmeans(Model.final, pairwise ~ CancerStage|Smoking, type="response", inf=TRUE)
```

### 8. Compute predicted probabilities of remission in lung for different patient characteristics attended by ID.doctor=2 in ID.hospital=1.

<!-- Predictions by cancer stage for non-smokers evaluated at mean `IL6`, `CRP` and `LenghOfStay` values: -->

```{r, message=FALSE, include=FALSE}
data.pred <- data.frame(ID.doctor=2, ID.hospital=1, IL6=mean(patients$IL6),
                        CRP=mean(patients$CRP), LengthofStay=mean(patients$LengthofStay),
                        CancerStage=c("I","II","III","IV"), Smoking="never")
data.pred$prob <- predict(Model.final, newdata=data.pred, type="response")
data.pred
```

<!-- Predictions by type of smoker for cancer stage I at mean `IL6`, `CRP` and `LenghOfStay` values: -->
```{r, message=FALSE, include=FALSE}
data.pred <- data.frame(ID.doctor=2, ID.hospital=1, IL6=mean(patients$IL6),
                        CRP=mean(patients$CRP), LengthofStay=mean(patients$LengthofStay),
                        CancerStage="I", Smoking=c("never","former","current"))
data.pred$prob <- predict(Model.final, newdata=data.pred, type="response")
data.pred
```

<!-- Predictions by IL6 for CancerStage="I" and Smoking="never": -->
```{r, message=FALSE, include=FALSE}
d <- 1:5
n.doctors <- length(d)
IL6.values <- as.numeric(quantile(patients$IL6,seq(0,1,0.05)))
n.IL6 <- length(IL6.values)

data.pred <- data.frame(ID.doctor=rep(d,each=n.IL6),
                        ID.hospital=1, IL6=rep(IL6.values,n.doctors),
                        CRP=mean(patients$CRP), LengthofStay=mean(patients$LengthofStay),
                        CancerStage="I", Smoking="never")

data.pred$prob <- predict(Model.final, newdata=data.pred, type="response")
head(data.pred)

ggplot(data.pred, aes(x=IL6, y=prob, group=ID.doctor, col=factor(ID.doctor))) +
  geom_line() +
  labs(x="IL6", y="Remission probabilities") + 
  ylim(0,1) + 
  ggtitle("CancerStage='I' and Smoking='never'")
```

<!-- Predictions by CRP for CancerStage="I" and Smoking="never": -->
```{r, message=FALSE, include=FALSE}
CRP.values <- as.numeric(quantile(patients$CRP,seq(0,1,0.05)))
n.CRP <- length(CRP.values)

data.pred <- data.frame(ID.doctor=rep(d,each=n.IL6),
                        ID.hospital=1, IL6=mean(patients$IL6), 
                        CRP=rep(CRP.values,n.doctors), LengthofStay=mean(patients$LengthofStay),
                        CancerStage="I", Smoking="never")

data.pred$prob <- predict(Model.final, newdata=data.pred, type="response")
head(data.pred)

ggplot(data.pred, aes(x=CRP, y=prob, group=ID.doctor, col=factor(ID.doctor))) +
  geom_line() +
  labs(x="CRP", y="Remission probabilities") + 
  ylim(0,1) + 
  ggtitle("CancerStage='I' and Smoking='never'") + 
  facet_wrap(~ CancerStage)
```
