---
title: "Chapter 2: Generalized Linear Mixed Models"
author: "Aritz Adin y Jaione Etxeberria"
date: "02-01-2024"
date-format: "MMM/YYYY"
format:
  html:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)  
```

# Exercise 1

Data from a clinical trial, involving 2 treatments (control and active drug), conducted at 11 randomly selected centres are discussed by [Beitler and Landis (1985)](https://www.jstor.org/stable/2530970). For the $i$-th center and $j$-th treatment, the proportion of $n_{ij}$ patients having a positive response ($y_{ij}$) is recorded below (*material extracted from Gallagher, 2023*):

| **Centre** | **Treatment** | $y_{ij}$ | $n_{ij}$ |
|:----------:|:-------------:|:--------:|:--------:|
|     1      |     drug      |    11    |    36    |
|     1      |    control    |    10    |    37    |
|     2      |     drug      |    16    |    20    |
|     2      |    control    |    22    |    32    |
|     3      |     drug      |    14    |    19    |
|     3      |    control    |    7     |    19    |
|  $\ldots$  |   $\ldots$    | $\ldots$ | $\ldots$ |

### 1. Load the data intored in the `multicentre.txt` file, convert `centre` variables to factor, and make a descriptive plot to visualize differences between treatments and/or centers.

```{r}
multicentre <- read.table(file="multicentre.txt", header=T, stringsAsFactors=T)
head(multicentre)

multicentre$centre <- as.factor(multicentre$centre)
str(multicentre)
```

We can make a descriptive plot of the data using the ggplot2 package

```{r}
library(ggplot2)

library(ggplot2)
ggplot(multicentre, aes(x=obs, y=centre, color=treat)) +
  geom_point() +
  labs(x="Observed probability", y="Centre")
```

We observe systematic differences between centres.

### 2. Which variable should be included as a fixed effect, and which as a random effect? Make a *design plot* to visually compare the magnitude of the effects of the `centre` and `treat` factors.

-   We want to compare these particular treatments (*experimental factor*), so we use fixed effects for the `treat` factor.

-   The eleven centres represents a random sample from the population about which we wish to make inferences (*random factor*), so we use random effects to model the `center` factor.

```{r, fig.cap="Average observed probabilities for each level of the factors `treat` and `center`"}
plot.design(obs ~ treat*centre, data=multicentre)
```

-   We see that the variability associated with the treatment is much lower than the variability associated with the centres.

-   We also see that the average probability of a positive response is higher in the *active drug* treatment.

### 3. Write the mixed logistic regression model equation.

$$y_{ij} | u_i  \sim Bin(n_{ij},\pi_{ij}), \quad i=1,\ldots,10 \quad \mbox{and} \quad j=1,2$$ $$\mbox{logit}(\pi_{ij}) = \beta_0 + \beta_1*x_{ij} + u_i, \quad u_i \sim N(0,\sigma_u^2)$$ where

-   $\pi_{ij}$ is the probability of a favourable outcome in a patient on the $i$-th centre and $j$-th treatment
-   $x_{ij}$ is an indicator variable for treatment (0=control, 1=active drug)
-   $u_i$ is a random variable associated with the $i$-th centre

### 4. Choose the correct structure for the random effects

```{r, message=FALSE, include=FALSE}
## Fit the models ##
library(lme4)

M0 <- glm(obs ~ 1 + treat, family="binomial", weights=n, data=multicentre)
M1 <- glmer(obs ~ 1 + treat + (1|centre), family="binomial", weights=n, data=multicentre)

## LRT for the variance component sigma2_u ##
LRT <- -2*(logLik(M0)-logLik(M1))
mean(1-pchisq(LRT,df=c(0,1)))

## We can also use the AIC/BIC to compare the models ##
data.frame(AIC=c(AIC(M0), AIC(M1)),
           BIC=c(BIC(M0), BIC(M1)),
           row.names=c("multicentre.glm","multicentre.glmer"))
```

### 5. Check if the treatment effect is significant.

We compare nested models (with and without `treat` factor) using the LRT with the `anova()` function

```{r, include=FALSE}
M1a <- glmer(obs ~ 1 + (1|centre), family="binomial", weights=n, data=multicentre)
M1b <- glmer(obs ~ 1 + treat + (1|centre), family="binomial", weights=n, data=multicentre)
anova(M1a,M1b)
```
<!-- **We conclude that the treatment effect is not statistically significant.** -->


### 6. Using the model with fixed and random effects, which is the probability of a positive response on the active drug group? Interpret the odds ratio $e^{\beta_1}$ and compute a 95% confidence interval.

```{r, include=FALSE}
library(emmeans)
emmeans(M1b, revpairwise~treat, type="response", inf=TRUE)
```

<!-- -  For $u_i=0$, the probability of a positive response over a population of centres under the active drug group ($x_{ij}=1$) is $\hat{\pi}_{ij}=0.376$ with a 95% confidence interval of $[0.220, 0.564]$. -->

<!-- -   The estimated odds of a patient showing a positive response on the active drug group relative to the control group is 1.08 with a 95% confidence interval of $[0.77, 1.51]$. -->


### 7. Using the best model, verify whether the model assumptions on the random effects are satisfied.

```{r, include=FALSE}
Model <- M1a

rand <- unlist(ranef(Model)$centre)
qqnorm(rand)
qqline(rand)

shapiro.test(rand)
```
<!-- We don't reject the null hypothesis of normality of the random effects. -->

### 8. Which is the estimated variance and 95% confidence interval of the within-centre random effect? Compute and interpret the intra-class correlation coefficient.

```{r, include=FALSE}
VarCorr(Model)
confint(Model, parm="theta_")
```

<!-- We can compute the intra-class correlation coefficient as $$\rho=\frac{\sigma^2_u}{\sigma^2_u+3.29}=1.169^2/(1.169^2+3.29) = 0.29347,$$ which means that 29% of the overall variability can be attributed to the differences between centres. -->

### 9. Compute the predicted probabilties of a positive response for each value of the dataset. Include those estimated probabilities in the descriptive plot of section 1).

```{r, fig.cap="Observed (dots) and predicted (asterisks) probabilities of a positive response for each centre"}
multicentre$pred <- predict(Model, newdata=multicentre, type="response")
head(multicentre)

ggplot(multicentre, aes(x=obs, y=centre, color=treat)) +
  geom_point() +
  geom_point(aes(x=pred, y=centre), shape=8, color="black") + 
  labs(x="Prob", y="Centre")
```

<!-- # Exercise 2 -->

<!-- See <https://stats.oarc.ucla.edu/r/dae/mixed-effects-logistic-regression/> -->

<!-- -   Several continuous and discrete covariates -->
<!-- -   Possibility to include 2 random effects -->
